A seamless, client-centric programming model for type safe web applications

### 2014-05-02 paper.tex
Address all language issues from R1.

Development of web applications is no task for the faint of heart.
The conventional method involves splitting your program into two logical parts,
writing the one in JavaScript, which is notorious even among its proponents for
being wonky and error-prone, and the other in any compiled or
server-interpreted language. Then, two are glued together using whichever
home-grown network protocol seems to fit the application. However, most web
applications are conceptually single entities, making this forced split an
undesirable hindrance which introduces new possibilities for defects, adds
development overhead and prevents code reuse.

Several attempts have been made at tackling this problem, but none is, in our
view, satisfactory. In this paper, we propose a functional programming model in
which a web application is written as a single program from which client and
server binaries are generated during compilation. Type annotations in the source
program control which parts are executed on the server and which are executed
on the client, and the two communicate using type safe RPC calls. Functions
which are not explicitly declared as server-side or client-side are usable by
either side.

@@

Code written in JavaScript, the only widely supported language for client-side
web applications, is often confusing and error-prone, much due to the
language's lack of modularity, encapsulation facilities and type safety.

@@

This code looks solid enough by web standards, but even this simple piece of
code contains no less than three asynchronous callbacks, two of which both read
and and modify the application's global state. This makes the program flow
non-obvious, and introduces unnecessary risk and complexity through the
haphazard state modifications.

@@

There are many conveivable improvements to the mainstream web development model
described in the previous section. We propose an alternative programming model
based on Haskell, in which web applications are written as a single program
rather than as two independent parts that just so happen to talk to each other.

@@

This is the entry point of our application. Recall that Haste.App applications
are written from the client's perspective, making this the starting point of
the client-side code.

@@

\subsection{Using server-side state}

While the Hello Server example illustrates how client-server communication is
handled, most web applications need to keep some server-side state as well.
At first glance, this would seem to be a trivial problem; can we not just
create our state holding elements in \lstinline!main! as usual?
Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.

@@

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{Server-side state: a misguided first attempt}
\label{lst:bad-state}
\end{listingfloat}

While this may at first seem to solve the problem of server-side state, this
approach raises a number of questions. Does \lstinline!visitorref! live on the
client or the server, or perhaps on both? What happens if our state holding
element is initialized in a non-deterministic way, and what if we assign it
a value on the server and then attempt to read that value on the client?
Recall that, as explained in section \ref{sec:helloserver}, the function
passed to \lstinline!runApp! should for all intents and purposes be regarded as
the program's entry point.

@@

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of a server lifted computation is never observable on the
client, the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server-side state.

@@

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{Server-side state: doing it properly}
\label{lst:good-state}
\end{listingfloat}

@@

The handshake is a simple operation; an \lstinline!MVar! is associated with the
connecting client's session identifier, and the pair is prepended to the
recipient list. Notice how the application's server state is passed in as the
function's first \lstinline!Useless RcptList! argument.

@@

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!export! & \parbox[t]{5cm}{Make a server-side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
a serializable argument.} \\
\lstinline!mkUseful! & \parbox[t]{5cm}{Extract the contents of a
\lstinline!Useless! value produced by \lstinline!liftServerIO! on the server.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

\paragraph{Two compilers}
The principal trick to our solution is compiling the same program twice; once
with a compiler that generates the server binary, and once with one that
generates JavaScript. Conditional compilation is used for a select few
functions, to enable slightly different behavior on the client and on the
server as necessary. As we believe that a monadic programming style is a good
fit for this application, we have chosen to use Haskell as the base of our
solution, which by default leads us to choose GHC as our server-side compiler.
For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
to provide the client-side code.

@@

As we can see, \lstinline!App! is a simple state monad, with underlying IO
capabilities to allow server-side computations to be forked from within it.
Its \lstinline!CallID! state element contains the identifier to be given to the
next exported function, and its other state element contains a mapping from
identifiers to exported functions.

What makes \lstinline!App! interesting is that computations in this monad are
executed on both the client and the server; once on server startup, and once
in the startup phase of each client. Its operations behave slightly differently
depending on whether they are executed on the client or on the server.
Execution is deterministic, ensuring that the same sequence of
\lstinline!CallID!s are generated during every run, both on the server and on
all clients. This is necessary to ensure that any particular call identifier
always refers to the same server-side function on all clients.

After all common code has been executed, the program flow diverges between the
client and the server; client-side, \lstinline!runClient! launches the
application's \lstinline!Client! computation whereas on the server, this
computation is discarded, and the server instead goes into an event loop,
waiting for calls from the client.

@@

The client-side definition of \lstinline!liftServerIO! is as simple as can be.

@@

Its argument is simply ignored, and an opaque dummy value is returned; recall
that server-lifted IO computations are not observable on the client. The
server-side implementation is only marginally more complex.

@@

This concept of client-side identifiers being sent to the server and used as
indices into a table mapping identifiers to remotely accessible functions is an
extension of the concept of ``static values'' introduced by Epstein et al with
Cloud Haskell\ \cite{cloudhaskell}, which is discussed further in section
\ref{sec:cloudhaskell}.

@@

The function corresponding to the RPC call's identifier is looked up in the
identifier-function mapping and applied to the received list of arguments;
the return value is bundled up with the a nonce, provided by the client to
match each return value to the RPC call that produced it since there may be
several RPC calls in progress at the same time, and sent back to the client.

@@

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both JavaScript and a format
        suitable for server-side execution as we want our web applications
        to be written and compiled as a single program.
  \item We want the language to provide decent support for a monadic
        programming style, as we our abstractions for cooperative multitasking
        and synchronous client-server communication are neatly expressible in
        this style.
\end{itemize}

@@

Like Conductance, Duetto suffers somewhat from its heritage: while the
client-side code is not memory-unsafe, as it is not possible to generate
memory-unsafe JavaScript code, its server-side counterpart unfortunately is.
Our reservations expressed about how network communication in Duetto can be
initiated implicitly apply to Duetto as well.

@@

Interactive multiplayer games is one type of application where this code
sharing may have a large impact. In order to ensure that players are not
cheating, a game server must keep track of the entire game state and send
updates to clients at regular intervals. However, due to network latency,
waiting for server input before rendering each and every frame is completely
impractical. Instead, the usual approach is to have each client continuously
compute the state of the game to the best of its knowledge, rectifying any
divergence from the game's ``official'' state whenever an update arrives from
the server. In this scenario, it is easy to see how reusing much of the same
game logic between the client and the server would be very important.

@@

This approach was inspired by Cloud Haskell\ \cite{cloudhaskell}, which
introduces the notion of ``static values''; values which are known at compile
time. Codifying this concept in the type system, to enable it to be used as a
basis for remote procedure calls, unfortunately requires some major changes to
the compiler. Cloud Haskell has a stopgap measure for unmodified compilers
wherein a remote table, pairing values with unique identifiers, are kept.
This explicit bookkeeping relies on the programmer to assign appropriate types
to both values themselves and their identifiers, breaking type safety.

@@

This approach also requires significant changes to the compiler, making it
unsuitable for our use case. Moreover, we are not entirely convinced about the
wisdom of allowing server-side execution of what is essentially arbitrary code
sent from the client which, in a web application context, is completely
untrustworthy. While analyzing code for improper behavior is certainly
possible, designing and enforcing a security policy sufficiently strict to
ensure correct behavior while flexible enough to be practically useful would be
an unwelcome burden on the programmer.

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.

\paragraph{Preventing divergence} While our solution completely safe from the
invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Recall that our solution does not modify
either of the compilers used. As such, both will produce code to execute all
computations called by \lstinline!main!. Consider the following example:

>>>>

Development of web applications is no task for the faint of heart.
The conventional method involves splitting your program into two logical parts,
writing the one in JavaScript, which is notorious even among its proponents for
being wonky and error-prone, and the other in any compiled or
server-interpreted language. Then, the two are glued together using whichever
home-grown network protocol seems to fit the application. However, most web
applications are conceptually single entities, making this forced split an
undesirable hindrance which introduces new possibilities for defects, adds
development overhead and prevents code reuse.

Several attempts have been made at tackling this problem, but none is, in our
view, satisfactory. In this paper, we propose a functional programming model in
which a web application is written as a single program from which client and
server binaries are generated during compilation. Type annotations in the source
program control which parts are executed on the server and which are executed
on the client, and the two communicate using type safe RPC calls. Functions
which are not explicitly declared as server side or client side are usable by
either side.

@@

Code written in JavaScript, the only widely supported language for client side
web applications, is often confusing and error-prone, much due to the
language's lack of modularity, encapsulation facilities and type safety.

@@

This code looks solid enough by web standards, but even this simple piece of
code contains no less than three asynchronous callbacks, two of which both read
and modify the application's global state. This makes the program flow
non-obvious, and introduces unnecessary risk and complexity through the
haphazard state modifications.

@@

There are many conceivable improvements to the mainstream web development model
described in the previous section. We propose an alternative programming model
based on Haskell, in which web applications are written as a single program
rather than as two independent parts that just so happen to talk to each other.

@@

This is the entry point of our application. Recall that Haste.App applications
are written from the client's perspective, making this the starting point of
the client side code.

@@

\subsection{Using server side state}

While the Hello Server example illustrates how client-server communication is
handled, most web applications need to keep some server side state as well.
At first glance, this would seem to be a trivial problem; can we not just
create our state holding elements in \lstinline!main! as usual?
Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.

@@

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{server side state: a misguided first attempt}
\label{lst:bad-state}
\end{listingfloat}

While this may at first seem to solve the problem of server side state, this
approach raises a number of questions. Does \lstinline!visitorref! live on the
client or the server, or perhaps on both? What happens if our state holding
element is initialized in a non-deterministic way, and what if we assign it
a value on the server and then attempt to read that value on the client?
Recall that, as explained in section \ref{sec:helloserver}, the function
passed to \lstinline!runApp! should for all intents and purposes be regarded as
the program's entry point.

@@

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of a server lifted computation is never observable on the
client, the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server side state.

@@

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{server side state: doing it properly}
\label{lst:good-state}
\end{listingfloat}

@@

The handshake is a simple operation; an \lstinline!MVar! is associated with the
connecting client's session identifier, and the pair is prepended to the
recipient list. Notice how the application's server state is passed in as the
function's argument.

@@

\begin{table}
\renewcommand{\arraystretch}{1.5}
\begin{center}
\begin{tabular}{|r|l|}
\hline
Function & Purpose \\
\hline
\lstinline!runApp! & Begin execution of a Haste.App program. \\
\lstinline!runClient! & \parbox[t]{5cm}{Lift a single \lstinline!Client!
computation into the \lstinline!App! monad. Must be at the very end of
an \lstinline!App! computation, which is enforced by the type system.} \\
\lstinline!liftServerIO! & \parbox[t]{5cm}{Lift an IO computation into the
\lstinline!App! monad. The computation and its result are exclusive to the
server, as enforced by the type system, and are not observable on the client.} \\
\lstinline!export! & \parbox[t]{5cm}{Make a server side function available to
be called remotely by the client.} \\
\lstinline!onServer! & \parbox[t]{5cm}{Dispatch a remote call to the server and
wait for its completion. The result of the remote computation is returned on
the client after it completes.} \\
\lstinline!<.>! & \parbox[t]{5cm}{Apply an \lstinline!export!ed function to
a serializable argument.} \\
\lstinline!mkUseful! & \parbox[t]{5cm}{Extract the contents of a
\lstinline!Useless! value produced by \lstinline!liftServerIO! on the server.} \\
\lstinline!getSessionID! & \parbox[t]{5cm}{Get the unique identifier for
the current session. This is a pure convenience function, to relieve
programmers of the burden of session bookkeeping.} \\
\hline
\end{tabular}
\end{center}
\caption{Core functions of Haste.App}
\label{tbl:api}
\end{table}

@@

\paragraph{Two compilers}
The principal trick to our solution is compiling the same program twice; once
with a compiler that generates the server binary, and once with one that
generates JavaScript. Conditional compilation is used for a select few
functions, to enable slightly different behavior on the client and on the
server as necessary. As we believe that a monadic programming style is a good
fit for this application, we have chosen to use Haskell as the base of our
solution, which by default leads us to choose GHC as our server side compiler.
For reasons outlined in section \ref{sec:haste}, the Haste compiler was chosen
to provide the client side code.

@@

As we can see, \lstinline!App! is a simple state monad, with underlying IO
capabilities to allow server side computations to be forked from within it.
Its \lstinline!CallID! state element contains the identifier to be given to the
next exported function, and its other state element contains a mapping from
identifiers to exported functions.

What makes \lstinline!App! interesting is that computations in this monad are
executed on both the client and the server; once on server startup, and once
in the startup phase of each client. Its operations behave slightly differently
depending on whether they are executed on the client or on the server.
Execution is deterministic, ensuring that the same sequence of
\lstinline!CallID!s are generated during every run, both on the server and on
all clients. This is necessary to ensure that any particular call identifier
always refers to the same server side function on all clients.

After all common code has been executed, the program flow diverges between the
client and the server; client side, \lstinline!runClient! launches the
application's \lstinline!Client! computation whereas on the server, this
computation is discarded, and the server instead goes into an event loop,
waiting for calls from the client.

@@

The client side definition of \lstinline!liftServerIO! is as simple as can be.

@@

Its argument is simply ignored, and an opaque dummy value is returned; recall
that server-lifted IO computations are not observable on the client. The
server side implementation is only marginally more complex.

@@

This concept of client side identifiers being sent to the server and used as
indices into a table mapping identifiers to remotely accessible functions is an
extension of the concept of ``static values'' introduced by Epstein et al with
Cloud Haskell\ \cite{cloudhaskell}, which is discussed further in section
\ref{sec:cloudhaskell}.

@@

The function corresponding to the RPC call's identifier is looked up in the
identifier-function mapping and applied to the received list of arguments.
The return value is paired with a nonce provided by the client to tie it to
its corresponding RPC call, since there may be several such calls in progress
at the same time. The pair is then sent back to the client.

@@

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both JavaScript and a format
        suitable for server side execution as we want our web applications
        to be written and compiled as a single program.
  \item We want the language to provide decent support for a monadic
        programming style, as our abstractions for cooperative multitasking
        and synchronous client-server communication are neatly expressible in
        this style.
\end{itemize}

@@

Like Conductance, Duetto suffers somewhat from its heritage: while the
client side code is not memory-unsafe, as it is not possible to generate
memory-unsafe JavaScript code, its server side counterpart unfortunately is.
Our reservations expressed about how network communication in Duetto can be
initiated implicitly apply to Duetto as well.

@@

Interactive multiplayer games are one type of application where this code
sharing may have a large impact. In order to ensure that players are not
cheating, a game server must keep track of the entire game state and send
updates to clients at regular intervals. However, due to network latency,
waiting for server input before rendering each and every frame is completely
impractical. Instead, the usual approach is to have each client continuously
compute the state of the game to the best of its knowledge, rectifying any
divergence from the game's ``official'' state whenever an update arrives from
the server. In this scenario, it is easy to see how reusing much of the same
game logic between the client and the server would be very important.

@@

This approach was inspired by Cloud Haskell\ \cite{cloudhaskell}, which
introduces the notion of ``static values''; values which are known at compile
time. Codifying this concept in the type system, to enable it to be used as a
basis for remote procedure calls, unfortunately requires some major changes to
the compiler. Cloud Haskell has a stopgap measure for unmodified compilers
wherein a remote table, pairing values with unique identifiers, is kept.
This explicit bookkeeping relies on the programmer to assign appropriate types
to both values themselves and their identifiers, breaking type safety.

@@

This approach also requires significant changes to the compiler, making it
unsuitable for our use case. Moreover, we are not entirely convinced about the
wisdom of allowing server side execution of what is essentially arbitrary code
sent from the client which, in a web application context, is completely
untrustworthy. While analyzing code for improper behavior is certainly
possible, designing and enforcing a security policy sufficiently strict to
ensure correct behavior while flexible enough to be practically useful would be
an unwelcome burden on the programmer.

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.

\paragraph{Preventing divergence} While our solution is completely safe from
the invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Recall that our solution does not modify
either of the compilers used. As such, both will produce code to execute all
computations called by \lstinline!main!. Consider the following example:
